---
title: "Practical Machine Learning - Course Project"
author: "Saptarsi Chowdhury"
date: "25 December, 2016"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.path='figures/',echo=TRUE, cache=TRUE, warning=FALSE, message=FALSE)
```
#  Weight Lifting Activity Recognition

## Executive Summary
Qualitative activity recognition differs from conventional activity recognition in a distinctive way. While the latter is concerned with recognising which activity is
performed, the former is concerned with assessing how (well) it is performed. The term `quality` is defined as 'conformance to specifications'. If the manner of execution of an activity is specified, then the quality can be measured by comparing its execution against this specification. Accelerometers were placed on the belt, arm-band and glove of 6 participants and on the dumbell to classify different exercises and count training repetitions. They performed barbell lifts correctly and incorrectly in 5 different ways.

The goal of this project is to apply machine learning techniques to accurately predict the manner in which the participants did the exercise. This report describes how the model for the project was built, its cross validation, expected out of sample error calculation, and the choices made. It was used successfully to accurately predict all 20 different test cases.

Source: [Weight Lifting Dataset](http://groupware.les.inf.puc-rio.br/har)

## Intended Results  
The goal of this project is to predict the manner in which they did the exercise. This is the "classe" variable in the training set. You may use any of the other variables to predict with. You should create a report describing how you built your model, how you used cross validation, what you think the expected out of sample error is, and why you made the choices you did. You will also use your prediction model to predict 20 different test cases. 

1. Your submission should consist of a link to a Github repo with your R markdown and compiled HTML file describing your analysis. Please constrain the text of the writeup to < 2000 words and the number of figures to be less than 5. It will make it easier for the graders if you submit a repo with a gh-pages branch so the HTML page can be viewed online (and you always want to make it easy on graders :-). 

2. You should also apply your machine learning algorithm to the 20 test cases available in the test data above. Please submit your predictions in appropriate format to the programming assignment for automated grading. See the programming assignment for additional details.  

## Data Preprocessing   
In order to reproduce the same results, you need a certain set of packages as well as setting a pseudo random seed equal to the one I have used.  
`Note`: To install, for instance, the `corrplot` package in R, run this command: `install.packages("corrplot")`.  
The following Libraries were used for this project, which you should install and load them in your working environment.  
```{r warning=FALSE, error=FALSE , cache = TRUE}
library(rattle)
library(caret)
library(rpart)
library(rpart.plot)
library(corrplot)
library(randomForest)
library(RColorBrewer)
```  
Finally, load the same seed with the following line of code:  
```{r warning=FALSE, error=FALSE , cache = TRUE}
set.seed(56789)
```  

## Getting Data  
First of all, set your current working directory.  
The following code fragment downloads the dataset to the `data` folder in the current working directory.  
```{r warning=FALSE, error=FALSE , cache = TRUE}
trainUrl <-"https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
testUrl <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"
trainFile <- "./data/pml-training.csv"
testFile  <- "./data/pml-testing.csv"
if (!file.exists("./data")) {
  dir.create("./data")
}
if (!file.exists(trainFile)) {
  download.file(trainUrl, destfile = trainFile, method = "curl")
}
if (!file.exists(testFile)) {
  download.file(testUrl, destfile = testFile, method = "curl")
}
rm(trainUrl)
rm(testUrl)
```  

## Reading Data  
After downloading the data from the data source, we can read the two csv files into two data frames.  
```{r warning=FALSE, error=FALSE , cache = TRUE}
trainRaw <- read.csv(trainFile)
testRaw <- read.csv(testFile)
dim(trainRaw)
dim(testRaw)
rm(trainFile)
rm(testFile)
```  
The training data set contains `r dim(trainRaw)[1]` observations and `r dim(trainRaw)[2]` variables, while the testing data set contains `r dim(testRaw)[1]` observations and `r dim(testRaw)[2]` variables. The `classe` variable in the training set is the outcome to predict.  

## Cleaning Data  
In this step, we will clean the dataset and get rid of observations with missing values as well as some meaningless variables.  

1. We clean the <b>Near Zero Variance</b> Variables.  
```{r warning=FALSE, error=FALSE , cache = TRUE}
NZV <- nearZeroVar(trainRaw, saveMetrics = TRUE)
head(NZV, 20)
training01 <- trainRaw[, !NZV$nzv]
testing01 <- testRaw[, !NZV$nzv]
dim(training01)
dim(testing01)
rm(trainRaw)
rm(testRaw)
rm(NZV)
```  

2. Removing some columns of the dataset that do not contribute much to the accelerometer measurements.  
```{r warning=FALSE, error=FALSE , cache = TRUE}
regex <- grepl("^X|timestamp|user_name", names(training01))
training <- training01[, !regex]
testing <- testing01[, !regex]
rm(regex)
rm(training01)
rm(testing01)
dim(training)
dim(testing)
```  

3. Removing columns that contain `NA's`.  
```{r warning=FALSE, error=FALSE , cache = TRUE}
cond <- (colSums(is.na(training)) == 0)
training <- training[, cond]
testing <- testing[, cond]
rm(cond)
```  

Now, the cleaned training data set contains `r dim(training)[1]` observations and `r dim(training)[2]` variables, while the testing data set contains `r dim(testing)[1]` observations and `r dim(testing)[2]` variables.  

Correlation Matrix of Columns in the Training Data set.  
```{r warning=FALSE, error=FALSE , cache = TRUE}
corrplot(cor(training[, -length(names(training))]), method = "color", tl.cex = 0.5)
```  

## Partitioning Training Set  
We split the cleaned training set into a pure training data set (70%) and a validation data set (30%). We will use the validation data set to conduct cross validation in future steps.  
```{r warning=FALSE, error=FALSE , cache = TRUE}
set.seed(56789) # For reproducibile purpose
inTrain <- createDataPartition(training$classe, p = 0.70, list = FALSE)
validation <- training[-inTrain, ]
training <- training[inTrain, ]
rm(inTrain)
```  
The Dataset now consists of `r dim(training)[2]` variables with the observations divided as following:  
1. Training Data: `r dim(training)[1]` observations.  
2. Validation Data: `r dim(validation)[1]` observations.  
3. Testing Data: `r dim(testing)[1]` observations.  

## Data Modelling  

### Decision Tree  
We fit a predictive model for activity recognition using <b>Decision Tree</b> algorithm.  
```{r warning=FALSE, error=FALSE , cache = TRUE}
modelTree <- rpart(classe ~ ., data = training, method = "class")
prp(modelTree)
```  

Now, we estimate the performance of the model on the <b>validation</b> data set.  
```{r warning=FALSE, error=FALSE , cache = TRUE}
predictTree <- predict(modelTree, validation, type = "class")
confusionMatrix(validation$classe, predictTree)
accuracy <- postResample(predictTree, validation$classe)
ose <- 1 - as.numeric(confusionMatrix(validation$classe, predictTree)$overall[1])
rm(predictTree)
rm(modelTree)
```  

The Estimated Accuracy of the Random Forest Model is `r accuracy[1]*100`% and the Estimated Out-of-Sample Error is `r ose*100`%.  

### Random Forest
We fit a predictive model for activity recognition using <b>Random Forest</b> algorithm because it automatically selects important variables and is robust to correlated covariates & outliers in general.  
We will use <b>5-fold cross validation</b> when applying the algorithm.  
```{r warning=FALSE, error=FALSE , cache = TRUE}
modelRF <- train(classe ~ ., data = training, method = "rf", trControl = trainControl(method = "cv", 5), ntree = 250)
modelRF
```  

Now, we estimate the performance of the model on the <b>validation</b> data set.  
```{r warning=FALSE, error=FALSE , cache = TRUE}
predictRF <- predict(modelRF, validation)
confusionMatrix(validation$classe, predictRF)
accuracy <- postResample(predictRF, validation$classe)
ose <- 1 - as.numeric(confusionMatrix(validation$classe, predictRF)$overall[1])
rm(predictRF)
```  

The Estimated Accuracy of the Random Forest Model is `r accuracy[1]*100`% and the Estimated Out-of-Sample Error is `r ose*100`%.  
Random Forests yielded better Results, as expected!  

## Predicting the manner of exercise for Test Data Set  
Now, we apply the <b>Random Forest</b> model to the original testing data set downloaded from the data source. We remove the problem_id column first.  
```{r warning=FALSE, error=FALSE , cache = TRUE}
rm(accuracy)
rm(ose)
predict(modelRF, testing[, -length(names(testing))])
```  

## Generating files to submit as answers for the assignment  
Function to generate files with predictions to submit for assignment.  
```{r warning=FALSE, error=FALSE , cache = TRUE}
pml_write_files = function(x){
  n = length(x)
  for(i in 1:n){
    filename = paste0("./Assignment_Solutions/problem_id_",i,".txt")
    write.table(x[i], file = filename, quote = FALSE, row.names = FALSE, col.names = FALSE)
  }
}
```  
Generating the Files.  
```{r warning=FALSE, error=FALSE , cache = TRUE}
pml_write_files(predict(modelRF, testing[, -length(names(testing))]))
rm(modelRF)
rm(training)
rm(testing)
rm(validation)
rm(pml_write_files)
```  